The Lasting Shadows of Economic Downturns

### 2024-02-15 outputs/paper/paper.qmd
Update paper

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-1
#| tbl-cap: Sample 1
kable(cleaned_four_states, format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("striped", "scale_down"), font_size = 7)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-2
#| tbl-cap: Sample 2
kable(head(cleaned_unemployment, 30), format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-3
#| tbl-cap: Sample 3
kable(cleaned_personal_income, format = "latex", booktabs = TRUE) |>
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
### the line graphs of average unemployment rate by four most populated states###
### which is something like simulated plot1###
```

```{r}
### the bar graph change of unemployment for four major states ###
### something like simulated plot2 ###
```

```{r}
### the line graph of personal income per capita by year ###
### something like simulated plot3 ###
```

```{r}
### something like simulated plot4 ###
```

```{r}
### something like simulated plot5 ###
```

@@

## Limitations and Next Steps
Due to limitations in accessing relevant datasets, this paper will not explore the diverse social impacts of recessions, specifically on birth rates and education, which were initially of primary interest. Drawing from Stuart's foundational research, we quantitatively assess the prolonged impacts of the 1980–1982 recession on individuals who were in their formative years—children, adolescents, and young adults—at the onset of this economic downturn. Stuart's findings reveal significant long-term detriments to educational attainment and income levels. Specifically, for individuals aged 0–10 in 1979, a 10 percent reduction in per capita earnings from 1979 to 1982 within their birth county correlates with a 4.7 percentage point (14.6 percent) reduction in the attainment of four-year college degrees, a decrease in earned income by $2,100 (5.2 percent), and a 1.6 percentage point (13.0 percent) increase in the likelihood of living in poverty during 2000–2013. The adverse effects on college graduation rates are notably pronounced and consistently significant for individuals aged 0–13 in 1979, and diminish to small and statistically non-significant levels for those aged 14–22. This pattern indicates that the recession's impact likely stems from a reduction in childhood human capital or a sustained decrease in parental financial capacity to support higher education. Future research will address the effects on education and birth rates once relevant and accessible datasets become available.

>>>>

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-1
#| tbl-cap: Sample 1
kable(cleaned_four_states, format = "latex", booktabs = TRUE, align = "c") |>
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), font_size = 7, full_width = FALSE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-2
#| tbl-cap: Sample 2
kable(head(cleaned_unemployment, 30), format = "latex", booktabs = TRUE, align = "c") |>
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), full_width = FALSE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-3
#| tbl-cap: Sample 3
kable(cleaned_personal_income, format = "latex", booktabs = TRUE, align = "c") |>
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), font_size = 7, full_width = FALSE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-4
#| tbl-cap: Sample 4
kable(head(cleaned_bussiness_pattern1, 10), format = "latex", booktabs = TRUE, align = "c") |>
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), full_width = FALSE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-5
#| tbl-cap: Sample 5
kable(head(cleaned_bussiness_pattern2, 10), format = "latex", booktabs = TRUE, align = "c") |>
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), full_width = FALSE)
```

@@

```{r fig.width=10, fig.height=5}
#| echo: false
#| message: false
#| warning: false
#| label: fig-1
#| fig-cap: plot 1
### the line graphs of average unemployment rate by four most populated states###
plot_data1 <-
  cleaned_four_states  |>
  select(year, state_name, average_unemployment_rate) |>
  arrange(year, state_name)

# Define custom colors for each state
state_colors <- c("California" = "red", "Florida" = "green", "New York" = "blue", "Texas" = "yellow")

# Create a line plot with custom styles
plot1 <- ggplot(plot_data1, aes(x = year, y = average_unemployment_rate, color = state_name)) +
  geom_line(size = 1) +
  scale_color_manual(values = state_colors) + # Set custom colors for lines
  geom_vline(xintercept = 1980, linetype="dashed", color = "red", size = 1) +
  geom_vline(xintercept = 1982, linetype="dashed", color = "red", size = 1) +
  annotate("text", x = 1980, y = max(plot_data1$average_unemployment_rate), label = "Recession Start", vjust = 2) +
  annotate("text", x = 1982, y = max(plot_data1$average_unemployment_rate), label = "Recession End", vjust = -0.5) +
  labs(title = "Average Unemployment Rate by Four Most Populated States and Year",
       x = "Year", y = "Average Unemployment Rate") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), # Center the plot title
        axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels (optional)
        legend.position = "bottom") # Move legend to the bottom

# Display the plot
print(plot1)

```

```{r fig.width=10, fig.height=5}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2
#| fig-cap: plot 2
### the bar graph change of unemployment for four major states ###
### something like simulated plot2 ###
plot_data2 <-
  cleaned_four_states  |>
  select(year, state_name, average_unemployment_rate) |>
  arrange(year, state_name) |>
  filter(year %in% c(1976, 1982, 1989))

# Define custom colors for the years
year_colors <- c("1976" = "green", "1982" = "red", "1989" = "blue")

# Create a bar plot with custom styles
plot2 <- ggplot(plot_data2, aes(x = state_name, y = average_unemployment_rate, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = year_colors) +  # Set custom colors for bars
  labs(title = "Change of Unemployment for Four Major States",
       x = "State", y = "Change in Unemployment Rate") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the plot title
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels (optional)
    legend.position = "bottom"  # Move legend to the bottom
  )

# Display the plot
print(plot2)
```

```{r fig.width=10, fig.height=5}
#| echo: false
#| message: false
#| warning: false
#| label: fig-3
#| fig-cap: plot 3
### the line graph of personal income per capita by year ###
plot_data3 <-
  cleaned_personal_income  |>
  filter(year %in% 1976:1989)

# Create a line plot with custom styles
plot3 <- ggplot(plot_data3, aes(x = year, y = `Per capita personal income (dollars) 2/`)) +
  geom_line(color = "forestgreen", size = 1) +  # Set line color and size
  labs(x = "Year", y = "Income Per Capita", 
       title = "Personal Income Per Capita by Year") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the plot title
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels (optional)
  )

# Display the plot
print(plot3)
```

```{r fig.width=10, fig.height=5}
#| echo: false
#| message: false
#| warning: false
#| label: fig-4
#| fig-cap: plot 4
plot_data4 <- cleaned_bussiness_pattern1 |>
  pivot_longer(
    cols = -year,
    names_to = "EmploymentType",
    values_to = "Count"
  ) |>
  filter(year %in% 1976:1989)

# Create a line plot
plot4 <- ggplot(plot_data4, aes(x = year, y = Count, color = EmploymentType)) +
  geom_line() +
  labs(
    title = "Business Pattern 1",
    x = "Year",
    y = "Count",
    color = "Employment Type"
  ) + theme_minimal()

# Display the plot
print(plot4)
```

```{r fig.width=10, fig.height=5}
#| echo: false
#| message: false
#| warning: false
#| label: fig-5
#| fig-cap: plot 5
plot_data5 <- cleaned_bussiness_pattern2 |>
  pivot_longer(
    cols = -year,
    names_to = "EmploymentType",
    values_to = "Count"
  ) |>
  filter(year %in% 1976:1989)

# Create a line plot
plot5 <- ggplot(plot_data5, aes(x = year, y = Count, color = EmploymentType)) +
  geom_line() +
  labs(
    title = "Business Pattern 1",
    x = "Year",
    y = "Count",
    color = "Employment Type"
  ) + theme_minimal()

# Display the plot
print(plot5)
```

@@

## Limitations and Next Steps

Due to limitations in accessing relevant datasets, this paper will not explore the diverse social impacts of recessions, specifically on birth rates and education, which were initially of primary interest. Drawing from Stuart's foundational research, we quantitatively assess the prolonged impacts of the 1980–1982 recession on individuals who were in their formative years—children, adolescents, and young adults—at the onset of this economic downturn. Stuart's findings reveal significant long-term detriments to educational attainment and income levels. Specifically, for individuals aged 0–10 in 1979, a 10 percent reduction in per capita earnings from 1979 to 1982 within their birth county correlates with a 4.7 percentage point (14.6 percent) reduction in the attainment of four-year college degrees, a decrease in earned income by \$2,100 (5.2 percent), and a 1.6 percentage point (13.0 percent) increase in the likelihood of living in poverty during 2000–2013. The adverse effects on college graduation rates are notably pronounced and consistently significant for individuals aged 0–13 in 1979, and diminish to small and statistically non-significant levels for those aged 14–22. This pattern indicates that the recession's impact likely stems from a reduction in childhood human capital or a sustained decrease in parental financial capacity to support higher education. Future research will address the effects on education and birth rates once relevant and accessible datasets become available.
